---
title: "Large-scale realistic simulations with `cashr`: Multiple $g_1$ and $\\pi_0$"
author: "Lei Sun"
date: "2018-12-13"
output:
  workflowr::wflow_html:
    code_folding: show
---

- This file will take ~30 hours to knit.

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
## load the data
r <- readRDS("../data/liver.rds")
```

```{r}
## load the script
source("../code/RNAseq_pipeline.R")
```

```{r}
## common simulation settings
ngene <- 1e4
nsamp <- 5
group <- rep(0 : 1, each = nsamp)
pi0.vec <- c(0.5, 0.9, 0.99)
q <- 0.1 # nominal FDR

## choose the 1e4 most expressed genes
Y = lcpm(r)
subset = top_genes_index(ngene, Y)
r = r[subset,]

## multiple testing methods
method.names <- c("cashr", "BH", "qvalue", "ashr", "locfdr")
method.col <- scales::hue_pal()(length(method.names))[c(5, 1, 2, 4, 3)]
```

```{r}
g1names <- c(
  "Gaussian",
  "Near Gaussian",
  "Spiky",
  "Skew",
  "Flat Top",
  "Bimodal"
)

sim.theta <- list(
  function (ngene, pi0) {
    ## Gaussian
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * ngene), 0, 2)
    ))
  },
  function (ngene, pi0) {
    ## near Gaussian
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * 0.6 * ngene)),
    rnorm(round((1 - pi0) * 0.4 * ngene), 0, 3)
    ))
  },
  function (ngene, pi0) {
    ## spiky
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * 0.4 * ngene), 0, 0.5),
    rnorm(round((1 - pi0) * 0.2 * ngene), 0, 2),
    rnorm(round((1 - pi0) * 0.4 * ngene), 0, 3)
    ))
  },
  function (ngene, pi0) {
    ## skew
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * 0.25 * ngene), -2, 2),
    rnorm(round((1 - pi0) * 0.25 * ngene), -1, 2),
    rnorm(round((1 - pi0) * 0.25 * ngene), 0, 1),
    rnorm(round((1 - pi0) * 0.25 * ngene), 1, 1)
    ))
  },
  function (ngene, pi0) {
    ## flat top
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * 0.5 * ngene), -1.5, 1.5),
    rnorm(round((1 - pi0) * 0.5 * ngene), 1.5, 1.5)
    ))
  },
  function (ngene, pi0) {
    ## bimodal
    sample(c(
    rep(0, round(pi0 * ngene)),
    rnorm(round((1 - pi0) * 0.5 * ngene), -1.5, 1),
    rnorm(round((1 - pi0) * 0.5 * ngene), 1.5, 1)
    ))
  }
)

XDP <- function (theta, qvalue, q) {
  disc <- (qvalue <= q)
  num.disc <- sum(disc)
  num.false.disc <- sum(theta[disc] == 0)
  num.true.disc <- sum(theta[disc] != 0)
  num.true <- sum(theta != 0)
  FDP <- num.false.disc / max(1, num.disc, na.rm = TRUE)
  TDP <- num.true.disc / num.true
  return(list(FDP = FDP, TDP = TDP))
}
```

```{r cache = TRUE}
nsim <- 3e3

set.seed(777)

sim.FDP <- sim.TDP <- c()

for (i in seq(sim.theta)) {
  for (j in seq(nsim)) {
    ## simulate correlated null z-scores
    counts <- r[, sample(ncol(r), 2 * nsamp)]
    summary <- count_to_summary(counts, group)
    Z <- summary$z
    sd.Z <- sd(Z)
    
    ## simulate and normalize standard deviations
    s <- summary$s
    s <- s / sqrt(mean(s^2))
    
    ## simulate normal means theta
    pi0 <- sample(pi0.vec, 1)
    theta <- sim.theta[[i]](ngene, pi0)
    
    ## construct observations
    X <- theta + s * Z
    
    ## test statistics
    z <- X / s
    p = pnorm(-abs(z)) * 2
    
    ## apply different methods
    fit.BH <- p.adjust(p, method = "BH")
    XDP.BH <- XDP(theta, fit.BH, q)

    fit.qvalue <- qvalue::qvalue(p)
    qvalue.qvalue <- fit.qvalue$qvalues
    XDP.qvalue <- XDP(theta, qvalue.qvalue, q)
    
    fit.locfdr <- tryCatch(locfdr::locfdr(z, plot = 0), error = function(e) {NA})
    qvalue.locfdr <- tryCatch(ashr::qval.from.lfdr(fit.locfdr$fdr), error = function(e) {rep(NA, ngene)})
    XDP.locfdr <- XDP(theta, qvalue.locfdr, q)

    fit.ashr <- ashr::ash(X, s, mixcompdist = "normal", method = "fdr")
    qvalue.ashr <- ashr::get_qvalue(fit.ashr)
    XDP.ashr <- XDP(theta, qvalue.ashr, q)
    
    fit.cashr <- cashr::cash(X, s)
    qvalue.cashr <- fit.cashr$qvalue
    XDP.cashr <- XDP(theta, qvalue.cashr, q)
    
    ## FDP aggregated
    sim.FDP <- rbind.data.frame(sim.FDP, data.frame(
      g1 = g1names[i],
      pi0 = pi0,
      sdZ = sd.Z,

      cashr = XDP.cashr$FDP,
      BH = XDP.BH$FDP,
      qvalue = XDP.qvalue$FDP,
      ashr = XDP.ashr$FDP,
      locfdr = XDP.locfdr$FDP
    ))
    
    ## TDP aggregated
    sim.TDP <- rbind.data.frame(sim.TDP, data.frame(
      g1 = g1names[i],
      pi0 = pi0,
      sdZ = sd.Z,

      cashr = XDP.cashr$TDP,
      BH = XDP.BH$TDP,
      qvalue = XDP.qvalue$TDP,
      ashr = XDP.ashr$TDP,
      locfdr = XDP.locfdr$TDP
    ))
  }
}
```

```{r}
saveRDS(sim.FDP, '../output/sim.FDP.rds')
saveRDS(sim.TDP, '../output/sim.TDP.rds')
```

```{r}
sim.FDP <- readRDS("../output/sim.FDP.rds")
sim.TDP <- readRDS("../output/sim.TDP.rds")
```

```{r}
library(ggplot2)

boxplot.quantile.sq <- function (x) {
  r <- sqrt(quantile(x^2, probs = c(0.05, 0.25, 0.5, 0.75, 0.95)))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  return(r)
}

mean.sq <- function (x) {
  r <- sqrt(mean(x^2))
  return(r)
}

mysqrt_trans <- function() {
  scales::trans_new("mysqrt", 
                    transform = base::sqrt,
                    inverse = function(x) ifelse(x<0, 0, x^2),
                    domain = c(0, Inf))
}
```

## Distribution of FDP

```{r}
FDP.ggdata <- reshape2::melt(sim.FDP, id.vars = c("g1", "pi0", "sdZ"), variable.name = "method", value.name = "FDP")
```

```{r}
FDP.plot <- ggplot(data = FDP.ggdata, aes(x = method, y = FDP, fill = method, color = method)) +
  stat_summary(fun.data = boxplot.quantile.sq, geom = "boxplot", position = "dodge", aes(width = 0.75), show.legend = FALSE) +
  stat_summary(fun.y = mean.sq, geom = "point", position = position_dodge(width = 0.9), show.legend = FALSE, shape = 13, size = 3) +
  scale_color_manual(labels = method.names, values = method.col) +
  scale_fill_manual(labels = method.names, values = alpha(method.col, 0.35)) +
  scale_y_continuous(trans = "mysqrt", breaks = c(0, 0.1, 0.2, 0.4, 0.6, 0.8)) +
  facet_grid(pi0 ~ g1, labeller = label_bquote(rows = pi[0] == .(pi0))) +
  scale_x_discrete(limits = rev(levels(FDP.ggdata$method))) +
  coord_flip() +
  geom_hline(yintercept = q, col = "black", linetype = "dashed", size = 1) +
  labs(y = "FDP") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_text(size = 15),
        axis.text.x = element_text(size = 12),
        strip.text = element_text(size = 15),
        legend.position = "bottom",
        legend.background = element_rect(color = "grey"),
        legend.text = element_text(size = 12))
```

## Density of $g_1$

```{r}
plotx <- seq(-6, 6, by = 0.01)

g1.density <- cbind.data.frame(
  plotx,
  dnorm(plotx, 0, 2),
  0.6 * dnorm(plotx) +
    0.4 * dnorm(plotx, 0, 3),
  0.4 * dnorm(plotx, 0, 0.5) +
    0.2 * dnorm(plotx, 0, 2) +
    0.4 * dnorm(plotx, 0, 3),
  0.25 * dnorm(plotx, -2, 2) +
    0.25 * dnorm(plotx, -1, 2) +
    0.25 * dnorm(plotx, 0, 1) +
    0.25 * dnorm(plotx, 1, 1),
  0.5 * dnorm(plotx, -1.5, 1.5) +
    0.5 * dnorm(plotx, 1.5, 1.5),
  0.5 * dnorm(plotx, -1.5, 1) +
    0.5 * dnorm(plotx, 1.5, 1)
)

colnames(g1.density)[-1] <- g1names

g1.ggdata <- reshape2::melt(g1.density,
                            id.vars = "plotx",
                            variable.name = "g1",
                            value.name = "ploty")
```

```{r}
g1.plot <- ggplot(data = g1.ggdata, aes(x = plotx, y = ploty)) +
  geom_line() +
  facet_wrap(~ g1, nrow = 1) +
  labs(x = expression(theta), y = expression(g[1](theta))) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 10),
        axis.title.y = element_text(size = 15),
        axis.text.y = element_text(size = 10),
        strip.text = element_text(size = 15),
        legend.position = "none",
        legend.background = element_rect(color = "grey"),
        legend.text = element_text(size = 12))
```

## Root MSE of FDP from nominal FDR

```{r}
sqrtMSE.ggdata <- aggregate(FDP ~ method + pi0 + g1, data = FDP.ggdata, FUN = function (x, q) {sqrt(mean((x - q)^2))}, q = q)

names(sqrtMSE.ggdata)[names(sqrtMSE.ggdata) == "FDP"] = 'rootMSE'
```

```{r}
sqrtMSE.plot <- ggplot(data = sqrtMSE.ggdata,
                       aes(x = factor(pi0),
                           y = rootMSE,
                           group = method,
                           col = method,
                           linetype = method,
                           shape = method)) +
  geom_line(size = 1.25) +
  geom_point(size = 3) + 
  scale_color_manual(values = method.col) +
  facet_wrap(~ g1, nrow = 1) +
  expand_limits(y = 0) +
  labs(y = 'Root MSE', x = expression(pi[0])) +
  theme(
    axis.title.y = element_text(size = 15),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 15),
    axis.text.x = element_text(size = 12),
    strip.text = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_blank()
  )
```

## Distribution of TDP

```{r}
TDP.ggdata <- reshape2::melt(sim.TDP, id.vars = c("g1", "pi0", "sdZ"), variable.name = "method", value.name = "TDP")

meanTDP.ggdata <- aggregate(TDP ~ method + pi0 + g1, data = TDP.ggdata, FUN = "mean")

names(meanTDP.ggdata)[names(meanTDP.ggdata) == "TDP"] = 'meanTDP'
```

```{r}
meanTDP.plot <- ggplot(data = meanTDP.ggdata,
                       aes(x = factor(pi0),
                           y = meanTDP,
                           group = method,
                           col = method,
                           linetype = method,
                           shape = method)) +
  geom_line(size = 1.25) +
  geom_point(size = 3) + 
  scale_color_manual(values = method.col) +
  facet_wrap(~ g1, nrow = 1) +
  expand_limits(y = 0) +
  labs(y = 'Mean TDP', x = expression(pi[0])) +
  theme(
    axis.title.y = element_text(size = 15),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 15),
    axis.text.x = element_text(size = 12),
    strip.text = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_blank()
  )
```

## Put together

```{r}
FDP_rootMSE_meanTDP_g1_pi0.plot <- gridExtra::arrangeGrob(
  g1.plot + 
    labs(title = expression(paste('(a): Distribution ', g[1], ' of non-null effects'))) +
    theme(plot.margin = grid::unit(c(20, 92, 20, 10), "points")) +
    theme(plot.title = element_text(size = 15)),

  FDP.plot +
    labs(title = bquote(paste('(b): Distribution of FDP'))) +
    theme(strip.text.x = element_blank(),
          plot.title = element_text(size = 15),
          plot.margin = grid::unit(c(5.5, 67.5, 5.5, 5.5), 'points')
    ),
  
  sqrtMSE.plot +
    labs(title = '(c): Root MSE of FDP from nominal FDR') +
    theme(strip.text = element_blank(),
          plot.title = element_text(size = 15),
          plot.margin = grid::unit(c(5.5, 5.5, 5.5, 15), 'points')
    ),

  meanTDP.plot +
    labs(title = '(d): Mean of TDP') +
    theme(strip.text = element_blank(),
          plot.title = element_text(size = 15),
          plot.margin = grid::unit(c(5.5, 5.5, 5.5, 15), 'points')
    ),

  heights = c(1.7, 2, 1.5, 1.5),

  top = grid::textGrob(bquote(paste("Nominal FDR = ", .(q))), gp = grid::gpar(fontsize = 17), hjust = 3.6)
)

ggsave("../figures/FDP_sqrtMSE_meanTDP_g1_pi0.pdf", FDP_rootMSE_meanTDP_g1_pi0.plot, height = 13.5, width = 15)
```

## Disclaimer

The RNA-seq data set used in this file for realistic simulations was created from the real human liver tissue RNA-seq gene expression read counts. In particular, the sample and gene identifiers have been removed from the data.

The RNA-seq gene expression data were originally generated by the GTEx Project, which was supported by the Common Fund of the Office of the Director of the National Institutes of Health, and by NCI, NHGRI, NHLBI, NIDA, NIMH, and NINDS. The data used for the analyses described in this paper were obtained from the GTEx Portal at [the GTEx Portal](https://www.gtexportal.org).
